---
title: "STAT778 Activity 9"
author: "Michael Tritle"
format: html
---

## Activity 2

### Weibull Generation by the Inverse CDF Method

Let $U \sim \mathrm{Uniform}(0,1)$. Using the Weibull CDF,

$F(x; k, \lambda) = 1 - e^{-(x/\lambda)^k}, \qquad x \geq 0,$

set

$U = F(X) = 1 - e^{-(X/\lambda)^k}.$

Then

$1-U = e^{-(X/\lambda)^k},$

$\log(1-U) = -(X/\lambda)^k,$

$-\log(1-U) = (X/\lambda)^k,$

$X = \lambda \big(-\log(1-U)\big)^{1/k}.$

Since $1-U \sim \mathrm{Uniform}(0,1)$ as well, this is equivalently

$X = \lambda (-\log U)^{1/k}.$

The following R function generates samples from a Weibull distribution via cdf inversion

```{r}
rweibull <- function(n, k, lambda) {
  if(n <= 0 || length(n) != 1) {
    stop("n must be a positive integer.")
  }
  if(k <= 0 || lambda <= 0) {
    stop("k and lambda must both be positive.")
  }
  
  u <- runif(n)
  x <- lambda * (-log(u))^(1 / k)
  x
}
```


## Activity 3

### Gompertz Generation by Generalized Rejection Sampling

We want to sample from

$f(x;\eta,\sigma) = \eta \sigma \exp\{\eta + \sigma x - \eta e^{\sigma x}\}, \qquad x \geq 0,$

with $\eta > 0$ and $\sigma > 0$.

We choose an exponential proposal

$g_r(x) = r e^{-rx}, \qquad x \geq 0,$

where $r > 0$ is a tuning parameter.

To use generalized rejection sampling, we need

$f(x) \leq c_r g_r(x)$

for all $x \geq 0$. Consider

$\log\!\left(\frac{f(x)}{g_r(x)}\right)
= \log\!\left(\frac{\eta \sigma}{r}\right) + \eta + (\sigma + r)x - \eta e^{\sigma x}.$

Differentiate with respect to $x$:

$\frac{d}{dx}\log\!\left(\frac{f(x)}{g_r(x)}\right)
= (\sigma + r) - \eta \sigma e^{\sigma x}.$

So the maximizer is

$x_r^\star = \max\left\{0,\; \frac{1}{\sigma}\log\!\left(\frac{\sigma + r}{\eta \sigma}\right)\right\}.$

Hence

$c_r = \frac{f(x_r^\star)}{g_r(x_r^\star)}.$

The rejection sampler is then:

1. Generate $Y \sim \mathrm{Exp}(r)$ and $U \sim \mathrm{Uniform}(0,1)$.
2. Accept $X = Y$ if

$U \leq \frac{f(Y)}{c_r g_r(Y)}.$

Otherwise reject and repeat.

```{r}
logf_gompertz <- function(x, eta, sigma) {
  log(eta) + log(sigma) + eta + sigma * x - eta * exp(sigma * x)
}

logg_exp <- function(x, rate) {
  log(rate) - rate * x
}

logc_from_t <- function(t, eta) {
  xs <- max(0, log((1 + t) / eta))
  log(eta / t) + eta + (1 + t) * xs - eta * exp(xs)
}

tune_rate <- function(eta, sigma) {
  opt <- optimize(
    f = function(t) logc_from_t(t, eta),
    interval = c(1e-6, max(20, 10 * eta))
  )
  sigma * opt$minimum
}

rgompertz_reject <- function(n, eta, sigma, rate = NULL, return_stats = FALSE) {
  if(length(n) != 1 || n <= 0 || n != as.integer(n)) {
    stop("n must be a positive integer.")
  }
  if(eta <= 0 || sigma <= 0) {
    stop("eta and sigma must be positive.")
  }

  if(is.null(rate)) {
    rate <- tune_rate(eta, sigma)
  }

  x_star <- max(0, log((sigma + rate) / (eta * sigma)) / sigma)
  logc <- logf_gompertz(x_star, eta, sigma) - logg_exp(x_star, rate)
  cval <- exp(logc)

  out <- numeric(n)
  accepted <- 0
  proposals <- 0
  uniforms_used <- 0

  while(accepted < n) {
    u1 <- runif(1)
    y <- -log(u1) / rate
    uniforms_used <- uniforms_used + 1

    u2 <- runif(1)
    uniforms_used <- uniforms_used + 1
    proposals <- proposals + 1

    log_alpha <- logf_gompertz(y, eta, sigma) - logc - logg_exp(y, rate)

    if(log(u2) <= log_alpha) {
      accepted <- accepted + 1
      out[accepted] <- y
    }
  }

  if(return_stats) {
    return(list(
      x = out,
      rate = rate,
      c = cval,
      acceptance_prob = 1 / cval,
      proposals = proposals,
      accepted = accepted,
      uniforms_used = uniforms_used,
      uniforms_per_accept = uniforms_used / accepted
    ))
  }

  out
}
```

```{r}
set.seed(778)

params <- data.frame(
  eta   = c(0.5, 0.5, 1, 2),
  sigma = c(0.5, 1, 1, 1.5)
)

results <- vector("list", nrow(params))

for(i in seq_len(nrow(params))) {
  sim <- rgompertz_reject(
    n = 10000,
    eta = params$eta[i],
    sigma = params$sigma[i],
    return_stats = TRUE
  )

  results[[i]] <- data.frame(
    eta = params$eta[i],
    sigma = params$sigma[i],
    variance_estimate = var(sim$x),
    tuned_rate = sim$rate,
    theoretical_acceptance = sim$acceptance_prob,
    empirical_acceptance = sim$accepted / sim$proposals,
    uniforms_per_accept = sim$uniforms_per_accept
  )
}

do.call(rbind, results)
```

Using the generalized rejection sampler, I generated $N=10{,}000$ Gompertz random deviates for several choices of $(\eta,\sigma)$ and estimated the variance in each case.

| $\eta$ | $\sigma$ | Variance Estimate | Tuned Rate | Theoretical Acceptance | Empirical Acceptance | Uniforms per Accept |
|---|---:|---:|---:|---:|---:|---:|
| 0.5 | 0.5 | 1.33768671 | 0.3928043 | 0.5854037 | 0.5848637 | 3.4196 |
| 0.5 | 1.0 | 0.32835543 | 0.7856085 | 0.5854037 | 0.5866479 | 3.4092 |
| 1.0 | 1.0 | 0.18181704 | 1.2399820 | 0.7037060 | 0.7022472 | 2.8480 |
| 2.0 | 1.5 | 0.03599115 | 3.2556920 | 0.8118233 | 0.8117542 | 2.4638 |

Observations: The empirical acceptance probabilities are very close to the theoretical acceptance probabilities, which indicates that the rejection sampler is implemented correctly.

The efficiency of the sampler is determined by the acceptance probability. If the acceptance probability is $1/c$, then the expected number of proposals per accepted draw is $c$. Since each proposal uses two uniforms, one to generate the exponential proposal and one for the accept/reject step, the expected number of uniforms per accepted draw is

$2c = \frac{2}{\text{acceptance probability}}.$

This agrees with the simulation output. For example, when the acceptance probability is about $0.5854$, the expected number of uniforms per accepted draw is

$\frac{2}{0.5854} \approx 3.42,$

which matches the observed values.

To determine whether the efficiency depends on the parameters, compare the first two rows. These use the same $\eta=0.5$ but different $\sigma$ values, and the theoretical acceptance probability is identical in both cases. This shows that the efficiency does not depend on $\sigma$.

However, the acceptance probability changes as $\eta$ changes, so the efficiency does depend on $\eta$. In particular, larger $\eta$ led to higher acceptance probabilities and therefore fewer uniforms needed per accepted draw.

Thus, for the tuned generalized rejection sampler, the efficiency depends on $\eta$ but not on $\sigma$.

## Activity 4

For the ratio-of-uniforms method, use the unnormalized density

$h(x;\mu,\kappa) = e^{\kappa \cos(x-\mu)}, \qquad x \in (\mu-\pi,\mu+\pi].$

Then

$C_h = \{(u,v): 0 < u \le \sqrt{h(v/u;\mu,\kappa)}\}.$

Since $\max_x \sqrt{h(x;\mu,\kappa)} = e^{\kappa/2}$, we take

$x_u = e^{\kappa/2}.$

Because $\mu \in [\pi,2\pi]$, the support satisfies $x \ge 0$, so $v = ux \ge 0$, and a rectangle is obtained with

$C_1 = [0,x_u] \times [0,y_u],$

where

$y_u = \max_{x \in [\mu-\pi,\mu+\pi]} x e^{(\kappa/2)\cos(x-\mu)}.$

For the banded region, write

$v = ux = \mu u + u(x-\mu),$

so

$|v-\mu u| \le |x-\mu| e^{(\kappa/2)\cos(x-\mu)}.$

Thus we may take

$\beta_1 = \mu, \qquad
\beta_0 = \max_{t \in [0,\pi]} t e^{(\kappa/2)\cos t},$

which gives the band

$C_2 = \{(u,v): 0 < u \le x_u,\ \beta_1 u - \beta_0 \le v \le \beta_1 u + \beta_0\}.$

The sampler proposes uniformly from either $C_1$ or $C_2$, accepts when $(u,v) \in C_h$, and returns $x=v/u$. The attribute `efficiency` is the realized acceptance proportion.

```{r}
rvonmises <- function(n, mu, kappa, type = 1) {
  if (length(n) != 1 || n <= 0 || n != as.integer(n)) {
    stop("n must be a positive integer.")
  }
  if (length(mu) != 1 || mu < pi || mu > 2 * pi) {
    stop("mu must satisfy pi <= mu <= 2*pi.")
  }
  if (length(kappa) != 1 || kappa <= 0) {
    stop("kappa must be positive.")
  }
  if (!(type %in% c(1, 2))) {
    stop("type must be either 1 or 2.")
  }

  h <- function(x) {
    exp(kappa * cos(x - mu))
  }

  x_low <- mu - pi
  x_high <- mu + pi
  x_u <- exp(kappa / 2)

  if (type == 1) {
    y_l <- 0
    y_u <- -optimize(
      f = function(x) -x * exp((kappa / 2) * cos(x - mu)),
      interval = c(x_low, x_high)
    )$objective
  }

  if (type == 2) {
    beta_1 <- mu
    beta_0 <- -optimize(
      f = function(t) -t * exp((kappa / 2) * cos(t)),
      interval = c(0, pi)
    )$objective
  }

  out <- numeric(n)
  accepted <- 0
  proposals <- 0

  while (accepted < n) {
    if (type == 1) {
      u <- runif(1, 0, x_u)
      v <- runif(1, y_l, y_u)
    } else {
      u <- runif(1, 0, x_u)
      v <- beta_1 * u + runif(1, -beta_0, beta_0)
    }

    proposals <- proposals + 1
    x <- v / u

    if (x > x_low && x <= x_high && u^2 <= h(x)) {
      accepted <- accepted + 1
      out[accepted] <- x
    }
  }

  attr(out, "efficiency") <- accepted / proposals
  out
}
```

2 examples

```{r}
set.seed(778)

x1 <- rvonmises(n = 10000, mu = 1.5 * pi, kappa = 2, type = 1)
attr(x1, "efficiency")

x2 <- rvonmises(n = 10000, mu = 1.5 * pi, kappa = 2, type = 2)
attr(x2, "efficiency")
```